\input{../preamble.tex}

\specID{SPEC-SW-012}
\specTitle{Audit Chat --- rate-limited conversational interface over the finding store}
\specStatus{DRAFT}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{Context}
The audit chat (internally referred to as \texttt{sweep chat} in the codebase) exposes a
per-entity conversational surface over the finding store. A reviewer selects an entity, asks a
question, and gets a streamed answer backed by the entity's audited data and any previously
extracted findings. Because every turn incurs an LLM call with web search enabled, and because
the same reviewer tends to retype variants of the same question, the service must enforce
per-user rate limits, deduplicate repeat questions, and window the finding history so the prompt
stays within budget. This spec defines those three guardrails and their observable behaviour.

\section{Goals}
\begin{itemize}[leftmargin=1.2em]
  \item Enforce a per-user request rate limit on the chat endpoint; return a typed \texttt{rate\_limited} event when the cap is exceeded.
  \item Deduplicate semantically equivalent repeated questions inside a short window and serve the previous answer instead of calling the model again.
  \item Window the finding history injected into the system prompt: bounded by count, ordered most-recent-first.
  \item Guarantee message persistence ordering: the user turn is stored before the model is called; the assistant turn is stored after the stream completes.
  \item Fail open on repository errors: a streaming client always receives a terminal \texttt{done} event, even after an internal error.
\end{itemize}

\section{Non-Goals}
\begin{nongoal}
This spec does not cover: (a) the content of the system prompt or quality of model output,
(b) long-term chat archival, analytics, or export, (c) per-organisation or tiered rate limits
beyond a single per-user counter, (d) the schema of extracted findings (covered elsewhere),
(e) multi-user collaborative chat on the same entity.
\end{nongoal}

\section{Invariants}
\begin{invariant}
Rate limiting is per-user and time-bounded: at most \texttt{RATE\_LIMIT} chat requests per rolling
\texttt{RATE\_WINDOW\_SECS} window. A request beyond the cap is refused before any model call is dispatched.
\end{invariant}

\begin{invariant}
Deduplication is scoped to a single (\texttt{userId}, \texttt{entityId}) pair: a normalized user
message that matches a previous message within the last \texttt{DEDUP\_WINDOW\_SECS} returns the
cached prior assistant response, no new model call, and no rate-limit decrement beyond the
accounting of the duplicate.
\end{invariant}

\begin{invariant}
Finding history injected into the prompt is bounded: at most \texttt{MAX\_FINDINGS\_IN\_PROMPT}
findings, ordered by creation time descending. Older findings are omitted, never summarized.
\end{invariant}

\begin{invariant}
Chat message persistence is ordered around the model call: the user message is persisted
\emph{before} the stream is opened; the assistant message is persisted \emph{after} the stream
terminates normally. A stream that errors mid-flight persists no assistant message.
\end{invariant}

\begin{invariant}
Every invocation of \texttt{chat} yields exactly one terminal event of type \texttt{done} or
\texttt{error}, even if the underlying repository or model call raises. No client is left hanging.
\end{invariant}

\section{Interface}
Public surface consumed by the transport layer.

\begin{lstlisting}[language=TypeScript]
export interface ChatMessage {
  role: "user" | "assistant";
  content: string;
  timestamp: string;        // ISO 8601
}

export interface Finding {
  id: string;
  entityId: string;
  category: string;
  title: string;
  summary: string;
  confidence: number;       // 0..1
  createdAt: string;
}

export type ChatEvent =
  | { type: "delta"; data: { text: string } }
  | { type: "finding"; data: Finding }
  | { type: "rate_limited"; data: { retryAfterSecs: number } }
  | { type: "duplicate"; data: { servedFromCache: true } }
  | { type: "error"; data: { error: string } }
  | { type: "done"; data: Record<string, never> };

export interface AuditChatRepository {
  checkRateLimit(userId: string): Promise<
    { allowed: true } | { allowed: false; retryAfterSecs: number }
  >;
  findDuplicate(
    userId: string,
    entityId: string,
    normalizedMessage: string,
  ): Promise<ChatMessage | null>;
  appendMessage(entityId: string, userId: string, msg: ChatMessage): Promise<void>;
  getHistory(entityId: string, userId: string): Promise<ChatMessage[]>;
  getFindings(entityId: string, limit?: number): Promise<Finding[]>;
}

export interface AuditChatService {
  chat(
    entityId: string,
    userId: string,
    message: string,
  ): AsyncGenerator<ChatEvent>;
}
\end{lstlisting}

\section{Scenarios}

\begin{scenario}{Nominal turn produces a streamed answer and stores both messages}
\Given user \texttt{U} has made no recent chat requests \\
\And entity \texttt{E} has $k < \texttt{MAX\_FINDINGS\_IN\_PROMPT}$ previous findings \\
\When \texttt{chat(E, U, "what changed last week?")} is iterated to completion \\
\Then the generator yields $\ge 1$ \texttt{delta} event, then a \texttt{done} event \\
\And \texttt{appendMessage} is called exactly twice: once with role \texttt{user} before the stream, once with role \texttt{assistant} after
\end{scenario}

\begin{scenario}{Rate limit exceeded}
\Given user \texttt{U} has already made \texttt{RATE\_LIMIT} successful chat requests in the current window \\
\When \texttt{chat(E, U, \ldots)} is invoked \\
\Then the generator yields a single \texttt{rate\_limited} event with a positive \texttt{retryAfterSecs} \\
\And then yields \texttt{done} \\
\And no model call is issued \\
\And no message is appended to history
\end{scenario}

\begin{scenario}{Duplicate question is served from cache}
\Given user \texttt{U} asked \texttt{"How is pricing evolving?"} on entity \texttt{E} $< \texttt{DEDUP\_WINDOW\_SECS}$ ago \\
\And the assistant answer was stored \\
\When \texttt{U} asks \texttt{"how is pricing evolving?"} again on \texttt{E} \\
\Then the service yields a \texttt{duplicate} event \\
\And yields the prior assistant content as one or more \texttt{delta} events \\
\And then yields \texttt{done} \\
\And no new model call is issued
\end{scenario}

\begin{scenario}{Finding history is windowed}
\Given entity \texttt{E} has 100 stored findings \\
\When a chat turn on \texttt{E} builds its system prompt \\
\Then the prompt embeds exactly \texttt{MAX\_FINDINGS\_IN\_PROMPT} findings \\
\And those findings are the \texttt{MAX\_FINDINGS\_IN\_PROMPT} with the most recent \texttt{createdAt}
\end{scenario}

\begin{scenario}{Mid-stream error is reported and terminated}
\Given the model provider raises partway through streaming \\
\When \texttt{chat} is iterated \\
\Then the generator yields zero or more \texttt{delta} events, then one \texttt{error} event, then a \texttt{done} event \\
\And no assistant message is persisted \\
\And the previously-persisted user message is not rolled back
\end{scenario}

\section{Acceptance Criteria}

\begin{ac}{AC-1}
Given a user with no recent chat activity, iterating \texttt{chat} to completion persists exactly one user message and one assistant message, in that order.
\end{ac}

\begin{ac}{AC-2}
Given a user at the rate-limit cap, \texttt{chat} yields a \texttt{rate\_limited} event followed by \texttt{done}, with no intermediate \texttt{delta} and no persisted message.
\end{ac}

\begin{ac}{AC-3}
Given a duplicate user message within \texttt{DEDUP\_WINDOW\_SECS}, \texttt{chat} yields a \texttt{duplicate} event and replays the cached assistant content without invoking the model.
\end{ac}

\begin{ac}{AC-4}
Message normalization for dedup is case-insensitive and whitespace-collapsing: \texttt{"Hello World"} and \texttt{"  hello  world  "} deduplicate to each other.
\end{ac}

\begin{ac}{AC-5}
When \texttt{getFindings(entityId)} returns more than \texttt{MAX\_FINDINGS\_IN\_PROMPT} results, the prompt builder includes only the \texttt{MAX\_FINDINGS\_IN\_PROMPT} most recent.
\end{ac}

\begin{ac}{AC-6}
Every terminating path of the generator --- nominal, rate-limited, duplicate, mid-stream error, repository error --- ends with exactly one \texttt{done} event.
\end{ac}

\begin{ac}{AC-7}
When the model stream throws mid-flight, the assistant message is \emph{not} persisted; the user message persisted before the stream remains.
\end{ac}

\begin{ac}{AC-8}
The rate-limit counter is incremented on successful request admission and is not incremented on a \texttt{duplicate} hit beyond its own accounting.
\end{ac}

\section{Traceability}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
AC & Test file & Test name \\
\midrule
AC-1 & \texttt{packages/sweep/tests/unit/audit-chat.spec.ts} & \texttt{persists user then assistant on nominal turn} \\
AC-2 & \texttt{packages/sweep/tests/unit/audit-chat.spec.ts} & \texttt{rate-limit short-circuits the stream} \\
AC-3 & \texttt{packages/sweep/tests/unit/audit-chat.spec.ts} & \texttt{duplicate replays cached answer} \\
AC-4 & \texttt{packages/sweep/tests/unit/dedup-normalize.spec.ts} & \texttt{normalization is case/whitespace insensitive} \\
AC-5 & \texttt{packages/sweep/tests/unit/audit-chat-prompt.spec.ts} & \texttt{finding window is bounded and most-recent-first} \\
AC-6 & \texttt{packages/sweep/tests/integration/audit-chat.spec.ts} & \texttt{every path yields a single done event} \\
AC-7 & \texttt{packages/sweep/tests/integration/audit-chat.spec.ts} & \texttt{mid-stream error does not persist assistant} \\
AC-8 & \texttt{packages/sweep/tests/unit/rate-limit.spec.ts} & \texttt{counter semantics across admission and dedup} \\
\bottomrule
\end{tabular}
\end{center}

\section{Open Questions}
\begin{itemize}[leftmargin=1.2em]
  \item Should rate limits be tiered (e.g. higher cap for admin roles), or is a single per-user cap sufficient for the interview scope?
  \item Is a pure string-equality dedup enough, or do we want embedding-based near-duplicate detection? Current draft is string-normalization only.
  \item Should duplicate hits still be stored in history (with an origin marker), or skipped entirely?
  \item What is the right default for \texttt{MAX\_FINDINGS\_IN\_PROMPT} (current proposal: 20)?
\end{itemize}

\end{document}
