\input{../preamble.tex}

\specID{SPEC-SW-011}
\specTitle{Editorial Copilot --- structured briefing drafts from audited data}
\specStatus{DRAFT}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{Context}
The sweep subsystem produces a stream of audited findings: each finding has been emitted by a
nano worker, classified by category, and (if a reviewer has processed it) either accepted,
rejected, or resolved. The Editorial Copilot turns that curated finding stream into structured
briefing drafts --- short, schema-validated summaries consumable by downstream editors or
dashboards. The point of this service is \emph{not} to be creative: it is to compose audited
inputs into a constrained shape, so that every briefing a human ever reads is traceable back to
an accepted finding, and never to raw model output.

\section{Goals}
\begin{itemize}[leftmargin=1.2em]
  \item Accept a set of finding references and return one briefing object conforming to a strict Zod schema.
  \item Reject any request that would ground the briefing on unaudited content --- only findings with a suitable audit state are eligible.
  \item Validate every model response against the output schema before returning; on failure, retry bounded times, then fail closed.
  \item Expose a deterministic \emph{provenance} mapping: every field in the output cites at least one source finding id.
  \item Run with a bounded token budget: inputs are truncated, outputs are length-capped per field.
\end{itemize}

\section{Non-Goals}
\begin{nongoal}
This spec does not cover: (a) free-form chat or conversational drafting (see SPEC-SW-012),
(b) publication, scheduling, or downstream distribution of briefings, (c) multilingual
translation, (d) the SEO scoring heuristic (owned by the client), (e) human-in-the-loop editing
after a briefing is produced, (f) persistence of briefings beyond returning them to the caller.
\end{nongoal}

\section{Invariants}
\begin{invariant}
No unaudited content is ever emitted. A briefing request whose finding set contains at least one
finding in state \texttt{unaudited} or \texttt{rejected} is refused with a typed error; no model
call is made.
\end{invariant}

\begin{invariant}
The briefing output is structured, not free-form: every response is parsed through the shared
Zod schema \texttt{BriefingSchema} and rejected on failure. The service returns a typed
\texttt{Briefing}, never a raw string.
\end{invariant}

\begin{invariant}
Provenance is total: for every top-level field of a returned briefing, the \texttt{provenance}
map contains at least one source finding id. A briefing with orphan fields is invalid.
\end{invariant}

\begin{invariant}
All text fields are length-bounded at schema level (title $\le$ 120 chars, summary $\le$ 400
chars, each bullet $\le$ 200 chars, $\le$ 6 bullets). Overruns are rejected, not truncated
silently.
\end{invariant}

\begin{invariant}
Model retries on schema-invalid output are bounded by \texttt{MAX\_REPAIR\_ATTEMPTS} (default
1). After the cap, the service returns a typed \texttt{BriefingError} --- never a partially
valid briefing.
\end{invariant}

\section{Interface}
Public surface; consumers include an admin briefing endpoint and internal aggregators.

\begin{lstlisting}[language=TypeScript]
import { z } from "zod";

export const BriefingSchema = z.object({
  title: z.string().min(8).max(120),
  summary: z.string().min(20).max(400),
  bullets: z.array(z.string().min(4).max(200)).min(1).max(6),
  tags: z.array(z.string().min(2).max(40)).max(8),
  category: z.enum([
    "trend", "anomaly", "quality", "coverage", "other",
  ]),
  // Field name -> non-empty list of source finding ids
  provenance: z.record(z.string(), z.array(z.string()).min(1)),
});

export type Briefing = z.infer<typeof BriefingSchema>;

export type BriefingError =
  | { kind: "unaudited_input"; findingIds: string[] }
  | { kind: "schema_invalid"; attempts: number; issues: string[] }
  | { kind: "model_unavailable"; cause: string };

export interface EditorialCopilot {
  draftBriefing(input: {
    findingIds: string[];
    angle?: string;      // optional user-provided framing
  }): Promise<
    | { ok: true; briefing: Briefing }
    | { ok: false; error: BriefingError }
  >;
}
\end{lstlisting}

\section{Scenarios}

\begin{scenario}{Nominal: audited findings produce a valid briefing}
\Given three finding ids all in state \texttt{accepted} \\
\When \texttt{draftBriefing} is invoked with those ids \\
\Then the response is \texttt{\{ ok: true, briefing \}} \\
\And \texttt{briefing} validates against \texttt{BriefingSchema} \\
\And every key of \texttt{briefing.provenance} maps to a non-empty subset of the input ids
\end{scenario}

\begin{scenario}{Refusal on unaudited input}
\Given one of the input finding ids is in state \texttt{unaudited} \\
\When \texttt{draftBriefing} is invoked \\
\Then the service returns \texttt{\{ ok: false, error: \{ kind: "unaudited\_input", findingIds: [\ldots] \} \}} \\
\And no call to the underlying model is made \\
\And the returned \texttt{findingIds} list contains the offending id
\end{scenario}

\begin{scenario}{Schema-invalid model output is repaired once then fails closed}
\Given the model returns a JSON object missing the \texttt{summary} field on attempt 1 \\
\And the model returns a JSON object missing the \texttt{bullets} field on attempt 2 \\
\When \texttt{draftBriefing} is invoked \\
\Then the service returns \texttt{\{ ok: false, error: \{ kind: "schema\_invalid", attempts: 2, issues: [\ldots] \} \}} \\
\And no partially valid briefing is emitted to the caller
\end{scenario}

\begin{scenario}{Orphan field is rejected at validation}
\Given the model returns a briefing with keys \texttt{title, summary, bullets, tags, category, provenance} \\
\And \texttt{provenance} omits the \texttt{summary} key \\
\When validation runs \\
\Then schema validation fails with an \texttt{orphan\_field} issue \\
\And the error propagates as \texttt{schema\_invalid}
\end{scenario}

\begin{scenario}{Model unavailable}
\Given the underlying cortex executor raises a network error \\
\When \texttt{draftBriefing} is invoked \\
\Then the service returns \texttt{\{ ok: false, error: \{ kind: "model\_unavailable", cause \} \}} \\
\And no exception escapes the service boundary
\end{scenario}

\section{Acceptance Criteria}

\begin{ac}{AC-1}
\texttt{draftBriefing} returns \texttt{ok: true} with a \texttt{Briefing} that parses successfully through \texttt{BriefingSchema} when all input findings are audited-accepted.
\end{ac}

\begin{ac}{AC-2}
\texttt{draftBriefing} returns \texttt{unaudited\_input} and performs no model call when at least one input id is \texttt{unaudited} or \texttt{rejected}.
\end{ac}

\begin{ac}{AC-3}
Every top-level field of a returned briefing appears as a key in \texttt{briefing.provenance} with a non-empty list of source ids, all drawn from the input finding ids.
\end{ac}

\begin{ac}{AC-4}
A model response exceeding any schema length bound (e.g. \texttt{summary} $>$ 400 chars) causes the response to be rejected; the caller sees \texttt{schema\_invalid}, never a truncated briefing.
\end{ac}

\begin{ac}{AC-5}
The number of retries on schema failure is capped at \texttt{MAX\_REPAIR\_ATTEMPTS}; beyond the cap the service returns \texttt{schema\_invalid} with the attempt count.
\end{ac}

\begin{ac}{AC-6}
When the cortex executor throws, \texttt{draftBriefing} returns \texttt{model\_unavailable}; no exception reaches the caller.
\end{ac}

\begin{ac}{AC-7}
The service never returns both \texttt{ok: true} and a briefing with an empty \texttt{bullets} array.
\end{ac}

\section{Traceability}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
AC & Test file & Test name \\
\midrule
AC-1 & \texttt{packages/sweep/tests/unit/editorial-copilot.spec.ts} & \texttt{happy path returns schema-valid briefing} \\
AC-2 & \texttt{packages/sweep/tests/unit/editorial-copilot.spec.ts} & \texttt{refuses unaudited findings without model call} \\
AC-3 & \texttt{packages/sweep/tests/unit/editorial-copilot.spec.ts} & \texttt{provenance covers every output field} \\
AC-4 & \texttt{packages/sweep/tests/unit/briefing-schema.spec.ts} & \texttt{rejects overlength fields} \\
AC-5 & \texttt{packages/sweep/tests/integration/editorial-copilot.spec.ts} & \texttt{schema repair retries are capped} \\
AC-6 & \texttt{packages/sweep/tests/integration/editorial-copilot.spec.ts} & \texttt{model failure returns model\_unavailable} \\
AC-7 & \texttt{packages/sweep/tests/unit/briefing-schema.spec.ts} & \texttt{bullets non-empty invariant} \\
\bottomrule
\end{tabular}
\end{center}

\section{Open Questions}
\begin{itemize}[leftmargin=1.2em]
  \item Should \texttt{category} be inferred from the dominant finding category, or selected freely by the model within the enum?
  \item Do we want a soft retry on \texttt{model\_unavailable} (with backoff), or leave retry policy to the caller?
  \item Should we expose a dry-run mode that returns only the prompt that would be sent, for audit purposes?
  \item How do we version \texttt{BriefingSchema}? A breaking field change affects all existing consumers.
\end{itemize}

\end{document}
