% SPEC-SW-003 — Resolution Service
\input{../preamble.tex}

\specID{SPEC-SW-003}
\specTitle{Resolution Service --- transactional write with audit trail}
\specStatus{DRAFT}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{Context}
Once a reviewer accepts a finding, the proposed change must be applied to the source of truth. This spec defines the \emph{writer}: it loads an accepted finding, validates its embedded action payload, resolves any ambiguous references (e.g. a name that matches multiple records), executes the mutations under a single transaction, records an audit row, and writes the outcome back to the pending buffer. Rejected or unaccepted findings are never written. Errors short-circuit into a visible failure state rather than a partial silent mutation.

\section{Goals}
\begin{itemize}[leftmargin=1.2em]
  \item Only operate on findings the reviewer has explicitly accepted.
  \item Validate every action payload against a strict schema before touching the DB.
  \item Surface ambiguity back to the reviewer (pending-selection state) instead of guessing.
  \item Apply every accepted resolution inside a single atomic transaction.
  \item Persist an audit row per resolution (who, what, when, rows touched, errors).
\end{itemize}

\section{Non-Goals}
\begin{nongoal}
This service does not decide acceptance (reviewer UI does). It does not generate findings (SPEC-SW-001) nor route notifications (SPEC-SW-002). It does not roll back previously committed resolutions.
\end{nongoal}

\section{Invariants}
\begin{invariant}
A resolution MUST NOT mutate the source of truth unless the finding's \texttt{accepted} flag is strictly \texttt{true}.
\end{invariant}
\begin{invariant}
All mutations for a single resolution happen inside one DB transaction. On any handler error, the transaction rolls back entirely; partial writes are impossible.
\end{invariant}
\begin{invariant}
Every resolution --- success, partial, or failed --- produces exactly one audit row with the outcome status, affected row count, and any error messages. The audit row is persistent (survives process restarts).
\end{invariant}
\begin{invariant}
If an action references an ambiguous target (e.g. a name that resolves to $>1$ candidate), the service returns \texttt{pending\_selection} without mutating the DB, and stores the selector options back in the pending buffer.
\end{invariant}

\section{Interface}

\begin{lstlisting}[language=TypeScript]
export type ResolutionStatus =
  | "success" | "partial" | "failed" | "pending_selection";

export interface ResolutionResult {
  status: ResolutionStatus;
  affectedRows: number;
  resolvedAt?: string;
  errors?: string[];
  pendingSelections?: PendingSelection[];
}

export class ResolutionService {
  constructor(
    entityRepo: EntityRepository,
    findingRepo: FindingRepository,
    auditSink: AuditSink,
    db: Database,
  );

  resolve(
    findingId: string,
    selections?: Record<string, string | number>,
  ): Promise<ResolutionResult>;
}
\end{lstlisting}

\section{Scenarios}

\begin{scenario}{Nominal accepted resolution}
\Given a finding with \texttt{accepted=true} and a valid \texttt{update\_field} action \\
\When \texttt{resolve(findingId)} is called \\
\Then the DB transaction opens, the field is updated, the audit row is inserted, the transaction commits \\
\And the returned \texttt{status} is \texttt{success} with \texttt{affectedRows > 0} \\
\And the pending buffer now shows the resolution outcome.
\end{scenario}

\begin{scenario}{Reviewer rejected (or never accepted)}
\Given a finding with \texttt{accepted=false} or \texttt{accepted=null} \\
\When \texttt{resolve(findingId)} is called \\
\Then no DB mutation occurs \\
\And the returned \texttt{status} is \texttt{failed} with an explanatory error \\
\And no audit row claims a successful resolution.
\end{scenario}

\begin{scenario}{Ambiguous reference --- first call}
\Given an accepted finding whose action references a name that matches two records \\
\When \texttt{resolve(findingId)} is called without \texttt{selections} \\
\Then no DB mutation occurs \\
\And the returned \texttt{status} is \texttt{pending\_selection} \\
\And the pending buffer is updated with the selector options so the reviewer can pick.
\end{scenario}

\begin{scenario}{Ambiguous reference --- second call with selection}
\Given the previous pending selection exists \\
\When \texttt{resolve(findingId, \{key: chosenId\})} is called \\
\Then the chosen id is used, the transaction runs, and \texttt{status} is \texttt{success}.
\end{scenario}

\begin{scenario}{Handler error rolls back}
\Given an accepted finding with two actions where the second throws \\
\When \texttt{resolve(findingId)} is called \\
\Then the entire transaction rolls back \\
\And \texttt{affectedRows = 0} \\
\And the audit row records \texttt{failed} with the error message.
\end{scenario}

\begin{scenario}{Malformed payload}
\Given an accepted finding whose \texttt{resolutionPayload} fails schema validation \\
\When \texttt{resolve(findingId)} is called \\
\Then no DB mutation occurs \\
\And the returned \texttt{status} is \texttt{failed} with a schema error \\
\And an audit row is still written with that failure.
\end{scenario}

\section{Acceptance Criteria}

\begin{ac}{AC-1}
Calling \texttt{resolve} on a finding with \texttt{accepted != true} produces \texttt{status=failed} and zero DB mutations.
\end{ac}
\begin{ac}{AC-2}
A malformed or schema-invalid \texttt{resolutionPayload} produces \texttt{status=failed} with a descriptive error and no DB mutations.
\end{ac}
\begin{ac}{AC-3}
All mutations for one \texttt{resolve} call execute inside a single transaction; if any handler throws, the transaction rolls back and \texttt{affectedRows=0}.
\end{ac}
\begin{ac}{AC-4}
An ambiguous reference returns \texttt{status=pending\_selection}, writes selector options to the pending buffer, and performs no mutation.
\end{ac}
\begin{ac}{AC-5}
When \texttt{selections} are provided and disambiguate the action, the transaction runs to completion and \texttt{status} is \texttt{success} or \texttt{partial}.
\end{ac}
\begin{ac}{AC-6}
Every call to \texttt{resolve} produces exactly one audit row, regardless of outcome (success, partial, failed, pending\_selection).
\end{ac}
\begin{ac}{AC-7}
The audit row contains: finding id, reviewer id, action kinds, affected row count, final status, error messages (if any), and a UTC timestamp.
\end{ac}
\begin{ac}{AC-8}
Concurrent \texttt{resolve} calls on the same finding are serialized; only the first commits, the second observes the resolved state and is a no-op with \texttt{status=failed} (\emph{already resolved}).
\end{ac}

\section{Traceability}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
AC & Test file & Test name \\
\midrule
AC-1 & \texttt{tests/unit/resolution.guards.spec.ts}       & \texttt{refuses unaccepted finding} \\
AC-2 & \texttt{tests/unit/resolution.payload.spec.ts}      & \texttt{rejects malformed payload} \\
AC-3 & \texttt{tests/integration/resolution.tx.spec.ts}    & \texttt{rolls back on handler error} \\
AC-4 & \texttt{tests/integration/resolution.ambiguous.spec.ts} & \texttt{returns pending\_selection} \\
AC-5 & \texttt{tests/integration/resolution.selection.spec.ts} & \texttt{commits with selection} \\
AC-6 & \texttt{tests/integration/resolution.audit.spec.ts} & \texttt{writes one audit row per call} \\
AC-7 & \texttt{tests/integration/resolution.audit.spec.ts} & \texttt{audit row contains full context} \\
AC-8 & \texttt{tests/integration/resolution.concurrency.spec.ts} & \texttt{serializes concurrent resolves} \\
\bottomrule
\end{tabular}
\end{center}

\section{Open Questions}
\begin{itemize}[leftmargin=1.2em]
  \item Audit sink: dedicated SQL table vs append-only log --- do we need queryable history from the reviewer UI?
  \item Should \texttt{partial} status be allowed at all, or should any handler error force \texttt{failed} + full rollback?
  \item Do we expose an \emph{undo} operation that compensates a prior resolution, or is rollback strictly at transaction scope?
\end{itemize}

\end{document}
