\input{../preamble.tex}

\specID{SPEC-SH-002}
\specTitle{AppError hierarchy --- structured error propagation}
\specStatus{DRAFT}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{Context}
Errors flowing out of controllers, services, and agent cortices must carry a stable, serialisable contract so that transport layers (HTTP, queue workers, log pipelines) can map them to status codes and structured logs without string-matching messages. The \texttt{AppError} hierarchy in \texttt{packages/shared/src/utils/error.ts} defines a small, closed set of error classes --- \texttt{AppError}, \texttt{ValidationError}, \texttt{NotFoundError}, \texttt{UnauthorizedError}, \texttt{SystemError} --- plus two helpers (\texttt{isAppError}, \texttt{getErrorMessage}) and a log formatter (\texttt{formatError}).

\section{Goals}
\begin{itemize}[leftmargin=1.2em]
  \item Provide a single base class (\texttt{AppError}) carrying a numeric \texttt{statusCode} for transport mapping.
  \item Offer four subclasses with fixed default status codes (400, 404, 401, 500) and distinct \texttt{name} values.
  \item Preserve \texttt{instanceof} relationships across the hierarchy so narrowing works predictably.
  \item Provide a type guard (\texttt{isAppError}) and pure helpers (\texttt{getErrorMessage}, \texttt{formatError}) that accept \texttt{unknown}.
\end{itemize}

\section{Non-Goals}
\begin{nongoal}
This hierarchy does not define transport-level serialisation (JSON body shape, HTTP headers), localisation of messages, or error codes beyond \texttt{statusCode}. It also does not aggregate multiple errors --- each instance describes one failure.
\end{nongoal}

\section{Invariants}
\begin{invariant}
Every subclass \texttt{C} of \texttt{AppError} satisfies: \texttt{(new C(msg)) instanceof AppError \&\& (new C(msg)) instanceof Error}.
\end{invariant}

\begin{invariant}
The \texttt{name} property of each instance equals the class name exactly: \texttt{"AppError"}, \texttt{"ValidationError"}, \texttt{"NotFoundError"}, \texttt{"UnauthorizedError"}, \texttt{"SystemError"}.
\end{invariant}

\begin{invariant}
Default status codes are fixed: \texttt{AppError} = 500, \texttt{ValidationError} = 400, \texttt{NotFoundError} = 404, \texttt{UnauthorizedError} = 401, \texttt{SystemError} = 500. \texttt{statusCode} is \texttt{readonly}.
\end{invariant}

\begin{invariant}
\texttt{isAppError}, \texttt{getErrorMessage}, and \texttt{formatError} are total over \texttt{unknown}: they never throw, regardless of input.
\end{invariant}

\section{Interface}

\begin{lstlisting}[language=TypeScript]
export class AppError extends Error {
  readonly statusCode: number;
  constructor(message: string, statusCode?: number);
}

export class ValidationError extends AppError {
  constructor(message: string); // statusCode = 400
}

export class NotFoundError extends AppError {
  constructor(message: string); // statusCode = 404
}

export class UnauthorizedError extends AppError {
  constructor(message?: string); // statusCode = 401
}

export class SystemError extends AppError {
  constructor(message: string); // statusCode = 500
}

export function isAppError(error: unknown): error is AppError;
export function getErrorMessage(error: unknown): string;
export function formatError(error: unknown): {
  name?: string;
  message: string;
  stack?: string;
};
\end{lstlisting}

\section{Scenarios}

\begin{scenario}{Subclass carries correct status and name}
\Given a new \texttt{ValidationError("bad input")} \\
\When its fields are read \\
\Then \texttt{statusCode === 400} \\
\And \texttt{name === "ValidationError"} \\
\And \texttt{message === "bad input"}.
\end{scenario}

\begin{scenario}{Instanceof hierarchy}
\Given a \texttt{NotFoundError} instance \texttt{e} \\
\When checked with \texttt{instanceof} \\
\Then \texttt{e instanceof NotFoundError} is \texttt{true} \\
\And \texttt{e instanceof AppError} is \texttt{true} \\
\And \texttt{e instanceof Error} is \texttt{true}.
\end{scenario}

\begin{scenario}{isAppError narrows unknown}
\Given an \texttt{unknown} value that is a \texttt{SystemError} instance \\
\When passed to \texttt{isAppError} \\
\Then the guard returns \texttt{true} \\
\And TypeScript narrows the value to \texttt{AppError} in the true branch.
\end{scenario}

\begin{scenario}{getErrorMessage on arbitrary input}
\Given the inputs \texttt{new Error("x")}, \texttt{"y"}, \texttt{42}, \texttt{null} \\
\When each is passed to \texttt{getErrorMessage} \\
\Then it returns \texttt{"x"}, \texttt{"y"}, \texttt{"42"}, \texttt{"null"} respectively \\
\And the function never throws.
\end{scenario}

\begin{scenario}{formatError shape}
\Given an \texttt{AppError("oops")} \\
\When passed to \texttt{formatError} \\
\Then the returned object has \texttt{name}, \texttt{message}, and \texttt{stack} string fields \\
\And \texttt{name === "AppError"}.
\end{scenario}

\section{Acceptance Criteria}

\begin{ac}{AC-1}
\texttt{new AppError("m").statusCode === 500} and \texttt{new AppError("m", 418).statusCode === 418}.
\end{ac}

\begin{ac}{AC-2}
Each subclass sets the expected default \texttt{statusCode}: \texttt{ValidationError} = 400, \texttt{NotFoundError} = 404, \texttt{UnauthorizedError} = 401, \texttt{SystemError} = 500.
\end{ac}

\begin{ac}{AC-3}
For every subclass \texttt{C}, \texttt{(new C(...)).name === "C"} (exact class-name string match).
\end{ac}

\begin{ac}{AC-4}
Every subclass instance satisfies \texttt{instance instanceof AppError \&\& instance instanceof Error}.
\end{ac}

\begin{ac}{AC-5}
\texttt{isAppError(x)} returns \texttt{true} iff \texttt{x} is an instance of \texttt{AppError} or a subclass; otherwise \texttt{false}. It never throws on any \texttt{unknown} input.
\end{ac}

\begin{ac}{AC-6}
\texttt{getErrorMessage(x)} returns \texttt{x.message} when \texttt{x instanceof Error}, otherwise \texttt{String(x)}, for all tested inputs including \texttt{undefined}, \texttt{null}, primitives, and objects.
\end{ac}

\begin{ac}{AC-7}
\texttt{formatError(x)} returns an object with a \texttt{message: string} field for all inputs; when \texttt{x instanceof Error}, the object additionally contains \texttt{name} and \texttt{stack} matching \texttt{x}.
\end{ac}

\begin{ac}{AC-8}
\texttt{UnauthorizedError} instantiated with no argument has \texttt{message === "Unauthorized"} and \texttt{statusCode === 401}.
\end{ac}

\section{Traceability}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
AC & Test file & Test name \\
\midrule
AC-1 & \texttt{packages/shared/tests/app-error.spec.ts} & \texttt{AppError default and override statusCode} \\
AC-2 & \texttt{packages/shared/tests/app-error.spec.ts} & \texttt{subclass default statusCodes} \\
AC-3 & \texttt{packages/shared/tests/app-error.spec.ts} & \texttt{name equals class name} \\
AC-4 & \texttt{packages/shared/tests/app-error.spec.ts} & \texttt{instanceof hierarchy holds} \\
AC-5 & \texttt{packages/shared/tests/app-error.spec.ts} & \texttt{isAppError narrows unknown} \\
AC-6 & \texttt{packages/shared/tests/app-error.spec.ts} & \texttt{getErrorMessage total over unknown} \\
AC-7 & \texttt{packages/shared/tests/app-error.spec.ts} & \texttt{formatError shape for Error and non-Error} \\
AC-8 & \texttt{packages/shared/tests/app-error.spec.ts} & \texttt{UnauthorizedError default message} \\
\bottomrule
\end{tabular}
\end{center}

\section{Open Questions}
\begin{itemize}[leftmargin=1.2em]
  \item Should \texttt{AppError} carry an optional \texttt{cause} field (ES2022) for error chaining, and should \texttt{formatError} surface it?
  \item Should a \texttt{ConflictError} (409) and \texttt{RateLimitError} (429) be added, or kept out to preserve the small, closed hierarchy?
  \item Should \texttt{formatError} redact \texttt{stack} in production environments, or is that the logger's responsibility?
\end{itemize}

\end{document}
