% SPEC-SH-003 --- CompletenessScorer

\input{../preamble.tex}

\specID{SPEC-SH-003}
\specTitle{CompletenessScorer --- adaptive weight normalization on partial records}
\specStatus{DRAFT}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{Context}
Downstream consumers (matchers, validators, dashboards) score the quality of a record against a set of weighted components (fields, signals, sub-scores). In practice, some components are routinely absent --- the source system did not emit them, the ingestion pipeline dropped them, or the producer is still backfilling. A naive weighted sum \(\sum w_i \cdot v_i\) penalises partial records even when the available components are strong, and makes scores from records with different coverage non-comparable. \texttt{CompletenessScorer} is the canonical utility that performs \emph{adaptive weight normalization}: weights are redistributed across the components that are actually present, so the score stays on a stable \([0, 1]\) scale regardless of coverage.

\section{Goals}
\begin{itemize}[leftmargin=1.2em]
  \item Produce a score in \([0, 1]\) that is invariant to the \emph{presence} pattern of components, depending only on the observed values and their relative weights.
  \item Report both the normalised score and the coverage ratio, so callers can filter low-coverage results without losing them.
  \item Be deterministic, pure, and allocation-light: usable inside hot paths of matchers and sweep workers.
  \item Fail loudly on malformed inputs (negative weights, non-finite values) rather than silently producing garbage.
\end{itemize}

\section{Non-Goals}
\begin{nongoal}
This spec does \emph{not} define which components are scored, their business meaning, or their default weights. Those are owned by the calling module (e.g.\ the matcher spec). This spec also does not perform imputation: missing components stay missing, they are never filled with a default value before scoring.
\end{nongoal}

\section{Invariants}
\begin{invariant}
For any input where at least one component is present with a finite value \(v_i \in [0, 1]\) and a strictly positive weight \(w_i\):
\[
\mathrm{score} \;=\; \frac{\sum_{i \in P} w_i \cdot v_i}{\sum_{i \in P} w_i} \in [0, 1]
\]
where \(P\) is the set of present components. If \(P = \emptyset\), the scorer returns \texttt{null} for the score and \(0\) for coverage; it never throws for empty input.
\end{invariant}

\begin{invariant}
Coverage is reported as \(\mathrm{coverage} = \sum_{i \in P} w_i \;/\; \sum_i w_i\), using the caller-declared full weight set as the denominator. Coverage is always in \([0, 1]\) and is independent of component values.
\end{invariant}

\section{Interface}

\begin{lstlisting}[language=TypeScript]
export interface ScorerComponent {
  readonly key: string;
  readonly weight: number;  // > 0
  readonly value: number | null | undefined;  // in [0, 1] when present
}

export interface CompletenessResult {
  readonly score: number | null;   // null iff coverage === 0
  readonly coverage: number;       // in [0, 1]
  readonly presentKeys: readonly string[];
  readonly missingKeys: readonly string[];
}

export function scoreCompleteness(
  components: readonly ScorerComponent[]
): CompletenessResult;
\end{lstlisting}

\section{Scenarios}

\begin{scenario}{Full coverage}
\Given three components with weights \((0.5, 0.3, 0.2)\) and values \((0.8, 0.6, 1.0)\) \\
\When \texttt{scoreCompleteness} is called \\
\Then \texttt{score} equals \(0.5 \cdot 0.8 + 0.3 \cdot 0.6 + 0.2 \cdot 1.0 = 0.78\) \\
\And \texttt{coverage} equals \(1.0\) \\
\And \texttt{missingKeys} is empty.
\end{scenario}

\begin{scenario}{Partial coverage, weights redistribute}
\Given the same weights, with values \((0.8, \texttt{null}, 1.0)\) \\
\When \texttt{scoreCompleteness} is called \\
\Then \texttt{score} equals \((0.5 \cdot 0.8 + 0.2 \cdot 1.0) / (0.5 + 0.2) = 0.8571\ldots\) \\
\And \texttt{coverage} equals \(0.7\) \\
\And \texttt{missingKeys} contains the key of the absent component.
\end{scenario}

\begin{scenario}{All components missing}
\Given three components whose values are all \texttt{null} or \texttt{undefined} \\
\When \texttt{scoreCompleteness} is called \\
\Then \texttt{score} is \texttt{null} \\
\And \texttt{coverage} is \(0\) \\
\And no exception is thrown.
\end{scenario}

\begin{scenario}{Rejects malformed weights}
\Given a component with weight \(-0.1\) or \texttt{NaN} \\
\When \texttt{scoreCompleteness} is called \\
\Then the function throws a \texttt{RangeError} referencing the offending \texttt{key}.
\end{scenario}

\begin{scenario}{Rejects out-of-range values}
\Given a component with a present value of \(1.4\) \\
\When \texttt{scoreCompleteness} is called \\
\Then the function throws a \texttt{RangeError} and does not return a partial result.
\end{scenario}

\section{Acceptance Criteria}

\begin{ac}{AC-1}
Full-coverage inputs yield \(\mathrm{score} = \sum w_i v_i / \sum w_i\) and \(\mathrm{coverage} = 1\).
\end{ac}

\begin{ac}{AC-2}
Partial-coverage inputs yield a score computed only over present components, with weights renormalised to their own sum.
\end{ac}

\begin{ac}{AC-3}
Empty-coverage inputs yield \texttt{score === null} and \texttt{coverage === 0}, without throwing.
\end{ac}

\begin{ac}{AC-4}
Any non-finite or negative weight causes a \texttt{RangeError} naming the offending component key.
\end{ac}

\begin{ac}{AC-5}
Any present value outside \([0, 1]\) causes a \texttt{RangeError} naming the offending component key.
\end{ac}

\begin{ac}{AC-6}
For the same input, \texttt{scoreCompleteness} returns bitwise-equal results across calls (determinism, no floating-point drift from iteration order).
\end{ac}

\begin{ac}{AC-7}
\texttt{presentKeys} and \texttt{missingKeys} partition the input keys exactly (no duplicates, no omissions).
\end{ac}

\section{Traceability}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
AC & Test file & Test name \\
\midrule
AC-1 & \texttt{packages/shared/tests/completeness-scorer.spec.ts} & \texttt{full coverage matches weighted sum} \\
AC-2 & \texttt{packages/shared/tests/completeness-scorer.spec.ts} & \texttt{partial coverage renormalises weights} \\
AC-3 & \texttt{packages/shared/tests/completeness-scorer.spec.ts} & \texttt{empty coverage returns null score} \\
AC-4 & \texttt{packages/shared/tests/completeness-scorer.spec.ts} & \texttt{rejects non-finite or negative weights} \\
AC-5 & \texttt{packages/shared/tests/completeness-scorer.spec.ts} & \texttt{rejects values outside [0,1]} \\
AC-6 & \texttt{packages/shared/tests/completeness-scorer.spec.ts} & \texttt{deterministic across repeated calls} \\
AC-7 & \texttt{packages/shared/tests/completeness-scorer.spec.ts} & \texttt{presentKeys and missingKeys partition input} \\
\bottomrule
\end{tabular}
\end{center}

\section{Open Questions}
\begin{itemize}[leftmargin=1.2em]
  \item Should \texttt{0} weights be tolerated (component declared but intentionally ignored) or rejected alongside negatives?
  \item Should the scorer expose a variant that accepts values in \([-1, 1]\) for signed signals, or is that a separate utility?
  \item How are duplicate keys handled --- first wins, last wins, or hard error?
\end{itemize}

\end{document}
