\input{../preamble.tex}

\specID{SPEC-SH-001}
\specTitle{tryAsync --- typed error tuple for async boundaries}
\specStatus{DRAFT}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{Context}
Asynchronous boundaries in agent cortices, transports, and tool executors must never crash the host process with an uncaught rejection. The \texttt{tryAsync} utility wraps a \texttt{Promise}-returning thunk and projects its outcome into a discriminated \texttt{Result<T, Error>} union, so callers can branch on \texttt{ok} instead of \texttt{try/catch}. This spec covers \texttt{tryAsync} and the related \texttt{allSettled} helper living in \texttt{packages/shared/src/utils/async-handler.ts}.

\section{Goals}
\begin{itemize}[leftmargin=1.2em]
  \item Convert any thrown value (\texttt{Error}, string, object, primitive) into a well-typed \texttt{Result<T, Error>}.
  \item Guarantee the returned Promise \emph{never rejects}, regardless of the inner thunk's behaviour.
  \item Preserve the original \texttt{Error} instance when one is thrown (no wrapping, no mutation).
  \item Provide a parallel variant (\texttt{allSettled}) that maps an array of thunks to an array of \texttt{Result} values, one per operation, with no short-circuit on failure.
\end{itemize}

\section{Non-Goals}
\begin{nongoal}
This utility does not perform retry, timeout, or cancellation. Those concerns are delegated to \texttt{retry} and \texttt{withTimeout}. It also does not log, report, or classify errors --- classification belongs to \texttt{AppError} (SPEC-SH-002).
\end{nongoal}

\section{Invariants}
\begin{invariant}
For every input thunk \texttt{fn}, the returned Promise resolves to exactly one of:
\texttt{\{ ok: true, value: T \}} or \texttt{\{ ok: false, error: Error \}}. It never rejects, never resolves to \texttt{undefined}, and never returns both branches.
\end{invariant}

\begin{invariant}
When the thunk throws a value \texttt{e}, the resulting \texttt{error} field is \texttt{e} itself when \texttt{e instanceof Error}, otherwise \texttt{new Error(String(e))}. The original stack trace is preserved in the first case.
\end{invariant}

\begin{invariant}
\texttt{allSettled} preserves input order: \texttt{results[i]} corresponds to \texttt{operations[i]} for every \texttt{i}, and its length equals the input length.
\end{invariant}

\section{Interface}

\begin{lstlisting}[language=TypeScript]
export type Result<T, E = Error> =
  | { ok: true; value: T }
  | { ok: false; error: E };

export function tryAsync<T>(
  fn: () => Promise<T>,
): Promise<Result<T, Error>>;

export function allSettled<T>(
  operations: (() => Promise<T>)[],
): Promise<Result<T, Error>[]>;
\end{lstlisting}

\section{Scenarios}

\begin{scenario}{Nominal resolution}
\Given a thunk that resolves to the value \texttt{42} \\
\When \texttt{tryAsync(fn)} is awaited \\
\Then the result is \texttt{\{ ok: true, value: 42 \}} \\
\And the returned Promise does not reject.
\end{scenario}

\begin{scenario}{Thrown Error instance}
\Given a thunk that throws \texttt{new RangeError("bad")} \\
\When \texttt{tryAsync(fn)} is awaited \\
\Then the result is \texttt{\{ ok: false, error \}} \\
\And \texttt{error} is the exact \texttt{RangeError} instance (reference equality) \\
\And \texttt{error.stack} is preserved.
\end{scenario}

\begin{scenario}{Thrown non-Error value}
\Given a thunk that throws the string \texttt{"boom"} \\
\When \texttt{tryAsync(fn)} is awaited \\
\Then the result is \texttt{\{ ok: false, error \}} \\
\And \texttt{error instanceof Error} is \texttt{true} \\
\And \texttt{error.message === "boom"}.
\end{scenario}

\begin{scenario}{Synchronous throw inside async thunk}
\Given a thunk whose body throws synchronously before any \texttt{await} \\
\When \texttt{tryAsync(fn)} is awaited \\
\Then the result is \texttt{\{ ok: false, error \}} \\
\And the returned Promise does not reject.
\end{scenario}

\begin{scenario}{Parallel mixed outcomes via allSettled}
\Given an array of three thunks: one resolving, one throwing \texttt{Error}, one throwing a number \\
\When \texttt{allSettled(ops)} is awaited \\
\Then the returned array has length 3 \\
\And entries preserve input order \\
\And each entry is a well-formed \texttt{Result<T, Error>}.
\end{scenario}

\section{Acceptance Criteria}

\begin{ac}{AC-1}
Given a thunk that resolves to \texttt{v}, \texttt{tryAsync(fn)} resolves to \texttt{\{ ok: true, value: v \}} and the returned Promise state is fulfilled.
\end{ac}

\begin{ac}{AC-2}
Given a thunk that rejects with an \texttt{Error} instance \texttt{e}, \texttt{tryAsync(fn)} resolves to \texttt{\{ ok: false, error \}} where \texttt{error === e}.
\end{ac}

\begin{ac}{AC-3}
Given a thunk that rejects with a non-Error value \texttt{x} (string, number, object), \texttt{tryAsync(fn)} resolves to \texttt{\{ ok: false, error \}} where \texttt{error instanceof Error \&\& error.message === String(x)}.
\end{ac}

\begin{ac}{AC-4}
The Promise returned by \texttt{tryAsync} never rejects; attaching only a \texttt{.then} handler is sufficient to observe every outcome.
\end{ac}

\begin{ac}{AC-5}
\texttt{allSettled} returns an array whose length equals \texttt{operations.length} and whose order matches the input.
\end{ac}

\begin{ac}{AC-6}
In \texttt{allSettled}, each element is a \texttt{Result<T, Error>} following AC-2/AC-3 normalization for that operation, independent of other operations' outcomes.
\end{ac}

\begin{ac}{AC-7}
\texttt{tryAsync} adds no properties to the thrown \texttt{Error} and does not mutate its \texttt{message}, \texttt{name}, or \texttt{stack} fields.
\end{ac}

\section{Traceability}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
AC & Test file & Test name \\
\midrule
AC-1 & \texttt{packages/shared/tests/try-async.spec.ts} & \texttt{resolves ok on success} \\
AC-2 & \texttt{packages/shared/tests/try-async.spec.ts} & \texttt{preserves thrown Error instance} \\
AC-3 & \texttt{packages/shared/tests/try-async.spec.ts} & \texttt{wraps non-Error throws} \\
AC-4 & \texttt{packages/shared/tests/try-async.spec.ts} & \texttt{never rejects} \\
AC-5 & \texttt{packages/shared/tests/try-async.spec.ts} & \texttt{allSettled preserves order and length} \\
AC-6 & \texttt{packages/shared/tests/try-async.spec.ts} & \texttt{allSettled normalizes each entry} \\
AC-7 & \texttt{packages/shared/tests/try-async.spec.ts} & \texttt{does not mutate Error fields} \\
\bottomrule
\end{tabular}
\end{center}

\section{Open Questions}
\begin{itemize}[leftmargin=1.2em]
  \item Should \texttt{tryAsync} expose an overload accepting a \texttt{Promise<T>} directly, mirroring \texttt{withTimeout}?
  \item Should a \texttt{Result} helper (\texttt{unwrap}, \texttt{map}, \texttt{mapErr}) be added, or is discriminated-union pattern matching sufficient for consumers?
\end{itemize}

\end{document}
