% SPEC-ARCH-01 — Ontology

\input{./preamble-arch.tex}

\specID{SPEC-ARCH-01}
\specTitle{Ontology --- shared vocabulary across Thalamus, Sweep and SSA}
\specStatus{OVERVIEW}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{Purpose}
Explicit vocabulary for both human readers and LLM consumers. Every term below is used consistently throughout the architecture documents.

\section{Glossary}

\begin{longtable}{@{}p{0.22\linewidth}p{0.72\linewidth}@{}}
\toprule
\textbf{Term} & \textbf{Definition} \\
\midrule
\endhead

\textbf{Cortex} &
A domain-specialized execution unit. Owns its tools, skill prompts, SQL helpers, and cost budget. Isolated blast radius. \\

\textbf{Skill} &
A markdown prompt file, versioned with the code. The ``binary'' the cortex runs. \\

\textbf{Nano worker} &
A cheap, fast model instance (e.g.\ \texttt{gpt-5.4-nano}) doing bounded retrieval or classification in a swarm. \\

\textbf{Swarm} &
Execution \emph{primitive}: fan-out of $K$ parallel bounded workers + a curator / aggregator collapse. The repo has \emph{three flavours} (retrieval, audit, counterfactual) --- see \texttt{SPEC-ARCH-08 three-swarms}. Not a subsystem. \\

\textbf{Fish} &
One instance inside a counterfactual (sim) swarm --- one \texttt{sim\_run} playing a perturbed variant of a scenario for $N$ turns. The swarm is the school; each fish swims its own perturbation. \\

\textbf{Curator} &
A stronger model (or pure aggregator) that deduplicates, ranks, and synthesizes worker / fish outputs into a single ranked payload. \\

\textbf{Sweep} &
\emph{Subsystem}: continuous DB audit + HITL loop (scan $\to$ finding $\to$ Redis $\to$ reviewer $\to$ resolution $\to$ audit $\to$ feedback). Uses swarms as a primitive. Not to be confused with the swarm itself. \\

\textbf{Finding} &
A structured observation (inconsistency, threat, conjunction) surfaced by the swarm, pending human review. \\

\textbf{Resolution} &
A transactional DB write triggered by an accepted finding. Always audited, always reversible. \\

\textbf{HITL} &
Human-in-the-loop. The reviewer is a first-class system component, not an exception path. \\

\textbf{Dual-stream} &
OSINT (low-confidence, high-volume) fused with Field (high-confidence, restricted). Confidence is never self-promoted. \\

\textbf{Source fetcher} &
A typed adapter behind the \texttt{SourceFetcher} interface. One per external system. Swappable, mockable. \\

\bottomrule
\end{longtable}

\end{document}
