% SPEC-ARCH-08 — Three swarms

\input{./preamble-arch.tex}

\specID{SPEC-ARCH-08}
\specTitle{Three swarms, three jobs --- don't confuse them with Sweep}
\specStatus{OVERVIEW}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{The word ``swarm'' does three things}

``Swarm'' is an execution \emph{primitive} (fan-out + collapse), and the repo uses it for three very different things. ``Sweep'' is a \emph{subsystem} that consumes two of them. The word sharing is unfortunate; the jobs are not the same.

\begin{longtable}{@{}p{0.17\linewidth}p{0.23\linewidth}p{0.19\linewidth}p{0.17\linewidth}p{0.18\linewidth}@{}}
\toprule
\textbf{Primitive} & \textbf{Implementation} & \textbf{What one worker is} & \textbf{Collapse step} & \textbf{What it produces} \\
\midrule
\endhead

\textbf{Retrieval swarm} &
\texttt{packages/thalamus/src/explorer/nano-swarm.ts} (Thalamus Explorer) &
One nano LLM call, one source / sub-query. Bounded tokens + wall-clock. &
\textbf{Curator} (stronger LLM): dedupe, rank, confidence-tag. &
Retrieval results that get normalized into \texttt{research\_finding} rows. \\

\textbf{Audit swarm} &
\texttt{packages/sweep/src/services/nano-sweep.service.ts} (Sweep NanoSweep scanner) &
One nano LLM call, one DB record (or one payload slice). Budgeted, rate-limited. &
\textbf{Finding routing} + Redis dedup + rate-limit per category. &
Pending findings that enter the Sweep reviewer UI. \\

\textbf{Counterfactual swarm} &
\texttt{packages/sweep/src/sim/swarm.service.ts} + \texttt{swarm-fish.worker} (Sweep sim engine) &
One \textbf{fish} --- a full \texttt{sim\_run} of up to $N$ turns, playing a \emph{perturbed} variant of the scenario (seed + \texttt{PerturbationSpec}). Each fish is an agent, not a single call. &
\textbf{Aggregator} (\texttt{aggregator.service.ts} / \texttt{aggregator-telemetry.ts}): modal outcome, quorum, divergence score. &
A single \texttt{sweep\_suggestion} when \texttt{modal.fraction} $\geq 0.5$ AND modal action $\in \{accept, maneuver\}$ (UC3), or $K$ per-scalar suggestions (telemetry UC). \\

\bottomrule
\end{longtable}

\subsection{The three swarms differ on three axes}

\begin{enumerate}[leftmargin=1.4em]
  \item \textbf{What gets perturbed.} Retrieval: the \emph{input query / source}. Audit: the \emph{DB record under scrutiny}. Counterfactual: the \emph{scenario itself} via \texttt{PerturbationSpec} --- different operator personas, different observation noise, different god-channel event timing.
  \item \textbf{What one worker returns.} Retrieval: text summary + citations. Audit: a candidate finding (one row). Counterfactual: a full trajectory of turns --- actions, rationales, observable summaries --- persisted to \texttt{sim\_turn}.
  \item \textbf{What the collapse means.} Retrieval ranks by \emph{relevance}. Audit ranks by \emph{dedup}. Counterfactual takes a \emph{modal vote} over outcomes and only emits a suggestion if enough fish agreed --- it's Monte-Carlo uncertainty quantification, not ranking.
\end{enumerate}

\subsection{Counterfactual use-case slots}

\begin{longtable}{@{}p{0.24\linewidth}p{0.42\linewidth}p{0.16\linewidth}p{0.14\linewidth}@{}}
\toprule
\textbf{Use-case slot} & \textbf{What a fish is simulating} & \textbf{Runner} & \textbf{Terminal condition} \\
\midrule
\endhead

\texttt{uc3\_conjunction} &
A mission operator receiving a conjunction event and deciding \texttt{accept} / \texttt{reject} / \texttt{maneuver} under a perturbed $P_c$ and radar-confirmation profile &
\texttt{turn-runner-sequential.ts} --- each turn is one action &
Fish ends early on \texttt{accept} or \texttt{reject} \\

\texttt{uc1\_operator\_behavior} &
Operator reactions to god-channel events (launches, advisories, press) over a horizon, no terminal action &
\texttt{turn-runner-dag.ts} --- agents act in parallel per turn &
Fish runs to \texttt{maxTurns} \\

\texttt{uc\_telemetry\_inference} &
A single agent infers 14 scalar telemetry values for one satellite from a perturbed prior &
DAG runner, 1 agent $\times$ 1 turn &
Terminal after turn 0 \\

\bottomrule
\end{longtable}

\begin{quote}
\textbf{Sweep is \emph{what} the workers are doing when the product is auditing the KB.} The three swarms are \emph{how} the work is parallelized. Sweep's UI surface consumes audit-swarm findings \emph{and} counterfactual-swarm suggestions; Thalamus consumes retrieval-swarm findings.
\end{quote}

\section{Read ``IUU maritime'' concretely --- swarm + Sweep in the same frame}

\textbf{The swarm (parallel workers):} for each AIS track showing a $\geq 12$ h transponder gap ending inside a protected EEZ in the last 24 h --- typically a few hundred candidates per region per day --- a nano worker is spawned. Each worker is bounded: one candidate gap, $\sim 15$ s wall-clock, $\sim 4$k tokens, one shot. Its job is to pull the SAR tile covering the gap's spatial cone (Sentinel-1 GRD, Global Fishing Watch dark-targets feed), check whether a vessel of the right length class was imaged during the gap, and cross-reference the flag and owner against the port-authority bulletin for the same week. Output: a candidate finding \texttt{\{ vesselMmsi, gapStart, gapEnd, sarMatchConfidence, portBulletinHit, suggestedNarrative \}} or a null.

\textbf{The curator (strong model, one call):} receives the few hundred worker outputs, dedupes by vessel + temporal window, merges overlapping gaps, ranks by corroboration strength (AIS gap + SAR hit + port bulletin = strong; AIS gap alone = weak), attaches source-class metadata on every edge. Emits a shortlist of ranked findings --- maybe 30 from 400 candidates.

\textbf{Sweep the subsystem takes over:} shortlist lands in Redis, deduped against the last 30 days of findings for the same vessel, rate-limited per EEZ. The coast-guard analyst UI (the reviewer) opens a finding: \emph{``vessel X (MMSI 311\dots, flag Panama) went dark 180 nm WNW of port P, SAR-detected at T+4 h inside the MPA, reappeared T+14 h 40 nm from original bearing, port-bulletin hit: crew change filed at port Q 72 h later''}. The analyst accepts / rejects / edits; on accept, \texttt{Resolution.service} opens a transaction --- writes the \texttt{InterdictionRecommendation}, writes the audit row (who, when, what, why), emits an accept signal back into the next swarm run's prompt for that EEZ. On reject, the rationale is captured and also feeds back.

\textbf{The loop:} the swarm drafts, the curator ranks, Sweep gates with a human, and the feedback signal calibrates tomorrow's swarm against this reviewer's precision bar. Zero autonomous action, every decision traceable --- same shape as SSA, same shape as threat intel, same code.

The takeaway: \textbf{the architecture never changes across domains}. The OSINT / Field confidence split, the swarm primitive, the curator, the Sweep subsystem with its HITL reviewer, the audit ledger, the guardrails in code --- all identical. Only the schema, the fetcher bundle, and the skill prompts move.

\end{document}
