\input{../preamble.tex}

\specID{SPEC-TH-012}
\specTitle{Curator --- dedup, ranking, and consolidation of swarm output}
\specStatus{DRAFT}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{Context}
The Nano-Swarm (SPEC-TH-010) produces a wide, noisy set of retrieval artifacts, some duplicated,
some off-topic, some already known to the system. Before any of these are injected into the feed or
promoted into a persistent source list, they must be scored for \emph{relevance} (do they answer the
research intent?) and \emph{novelty} (do they add information we do not already have?). The Curator
is the last stage of the Explorer pipeline and turns the raw artifact stream into a typed
consolidation decision stream.

\section{Goals}
\begin{itemize}[leftmargin=1.2em]
  \item Score each artifact on two orthogonal axes: relevance \([0,1]\) and novelty \([0,1]\).
  \item Produce a ternary decision per artifact --- \texttt{inject}, \texttt{promote}, or \texttt{discard} --- using deterministic thresholds.
  \item Assign every artifact a coarse category so downstream consumers can route it.
  \item Remain operational without the underlying LLM: a heuristic fallback preserves ordering semantics when the model call fails.
  \item Preserve input order and cardinality: exactly one \texttt{CuratedItem} out per input artifact.
\end{itemize}

\section{Non-Goals}
\begin{nongoal}
This spec does not cover: the feed-injection SQL, the source-promotion logic, cross-cycle novelty
memory (novelty is evaluated against what the LLM or heuristic can infer from the batch), or the
prompt wording itself (implementation detail).
\end{nongoal}

\section{Invariants}
\begin{invariant}
The Curator is total: \texttt{curate(articles)} returns a list of exactly \texttt{articles.length} items, in the same order.
\end{invariant}

\begin{invariant}
\texttt{relevanceScore} and \texttt{noveltyScore} are clamped to \([0,1]\) for every returned item,
including for heuristic-path items.
\end{invariant}

\begin{invariant}
The \texttt{action} field is one of exactly \{\texttt{"inject"}, \texttt{"promote"}, \texttt{"discard"}\};
\texttt{category} is one of exactly \{\texttt{"MARKET"}, \texttt{"REVIEWS"}, \texttt{"DROPS"}, \texttt{"DISCOVERY"}\}.
Any out-of-range value from the model is coerced to \texttt{"discard"} / \texttt{"DISCOVERY"} respectively.
\end{invariant}

\begin{invariant}
An LLM failure on one batch does not contaminate other batches: the Curator falls back to the
heuristic scorer for that batch only and continues.
\end{invariant}

\begin{invariant}
Given \(relevance \ge 0.7\) and \(novelty \ge 0.5\), the heuristic path yields \texttt{action = "inject"}
(or stronger: \texttt{"promote"} if relevance \(\ge 0.8\) and entity count \(\ge 5\)).
\end{invariant}

\section{Interface}
The Curator consumes the artifact type produced by SPEC-TH-010 and emits a decision envelope.

\begin{lstlisting}[language=TypeScript]
export interface CuratedItem {
  url: string;
  title: string;
  body: string;
  relevanceScore: number;   // [0,1]
  noveltyScore: number;     // [0,1]
  action: "inject" | "promote" | "discard";
  category: "MARKET" | "REVIEWS" | "DROPS" | "DISCOVERY";
  reason: string;           // <= 200 chars, 1 sentence
  entities: ExtractedEntities;
}

export class ExplorerCurator {
  curate(articles: RetrievalArtifact[]): Promise<CuratedItem[]>;
}
\end{lstlisting}

\section{Scenarios}

\begin{scenario}{Nominal batch scoring}
\Given 25 retrieval artifacts and a healthy LLM transport \\
\When \texttt{curate} is invoked \\
\Then the artifacts are split into batches of at most 10 \\
\And each batch is scored with one LLM call \\
\And the returned list has exactly 25 items, in the original input order \\
\And every score is clamped to \([0,1]\)
\end{scenario}

\begin{scenario}{LLM unavailable for one batch}
\Given three batches where the second one causes the LLM call to reject \\
\When \texttt{curate} is invoked \\
\Then batches 1 and 3 are scored by the LLM \\
\And batch 2 is scored by the heuristic fallback \\
\And the final list preserves the original input order and length
\end{scenario}

\begin{scenario}{Malformed model output}
\Given the LLM returns JSON where \texttt{action} is \texttt{"maybe"} and \texttt{category} is \texttt{"OTHER"} \\
\When the batch is parsed \\
\Then those fields are coerced to \texttt{"discard"} and \texttt{"DISCOVERY"} \\
\And the item is still returned (not dropped)
\end{scenario}

\begin{scenario}{Heuristic promotion}
\Given an artifact with rich wine entities (\(\ge 5\)), concrete data points, and body \(\ge 1500\) chars \\
\When the heuristic path runs on it \\
\Then \texttt{relevanceScore} is at least 0.8 \\
\And \texttt{action} is \texttt{"promote"}
\end{scenario}

\begin{scenario}{Empty input}
\Given an empty artifact list \\
\When \texttt{curate} is invoked \\
\Then the result is \texttt{[]} \\
\And no LLM call is made
\end{scenario}

\section{Acceptance Criteria}

\begin{ac}{AC-1}
For every input artifact there is exactly one output item with the same \texttt{url}, in the same position.
\end{ac}

\begin{ac}{AC-2}
\texttt{relevanceScore} and \texttt{noveltyScore} are in \([0,1]\) for every returned item.
\end{ac}

\begin{ac}{AC-3}
\texttt{action} and \texttt{category} always belong to their declared enumerations; unknown values from the model are coerced to \texttt{"discard"} / \texttt{"DISCOVERY"}.
\end{ac}

\begin{ac}{AC-4}
A forced LLM failure on one batch results in that batch being heuristic-scored; the other batches are unaffected.
\end{ac}

\begin{ac}{AC-5}
An input of size 0 returns \texttt{[]} and triggers zero LLM calls.
\end{ac}

\begin{ac}{AC-6}
For an artifact with \(\ge 5\) entities, \(\ge 1\) data point, and body length \(\ge 1500\), the heuristic path yields \texttt{action \(\in\) \{"inject","promote"\}}.
\end{ac}

\begin{ac}{AC-7}
\texttt{reason} is a non-empty string no longer than 200 characters for every item.
\end{ac}

\begin{ac}{AC-8}
Batches submitted to the LLM contain at most 10 artifacts; a mock LLM transport asserts this bound.
\end{ac}

\section{Traceability}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
AC & Test file & Test name \\
\midrule
AC-1 & \texttt{packages/thalamus/tests/curator.order.spec.ts} & \texttt{preserves input order and cardinality} \\
AC-2 & \texttt{packages/thalamus/tests/curator.clamp.spec.ts} & \texttt{clamps scores to [0,1]} \\
AC-3 & \texttt{packages/thalamus/tests/curator.enum.spec.ts} & \texttt{coerces unknown enum values} \\
AC-4 & \texttt{packages/thalamus/tests/curator.fallback.spec.ts} & \texttt{falls back to heuristic on LLM failure} \\
AC-5 & \texttt{packages/thalamus/tests/curator.empty.spec.ts} & \texttt{empty input returns empty output} \\
AC-6 & \texttt{packages/thalamus/tests/curator.heuristic.spec.ts} & \texttt{promotes entity-rich artifacts} \\
AC-7 & \texttt{packages/thalamus/tests/curator.reason.spec.ts} & \texttt{reason is always bounded} \\
AC-8 & \texttt{packages/thalamus/tests/curator.batching.spec.ts} & \texttt{batches at most 10 artifacts per LLM call} \\
\bottomrule
\end{tabular}
\end{center}

\section{Open Questions}
\begin{itemize}[leftmargin=1.2em]
  \item Should novelty be evaluated against a cross-cycle memory (e.g. prior URLs, prior entity fingerprints) rather than from the LLM alone?
  \item Should the \texttt{category} enumeration be pluggable per-domain, or remain fixed at the Explorer level?
  \item Should the heuristic thresholds (0.7 / 0.5 / 0.8) be configurable, or fixed as part of the contract?
\end{itemize}

\end{document}
