% SPEC-TH-041 — Field-Correlation Cortex
% Real-time OSINT x Field corroboration with sub-second latency on critical events.

\input{../preamble.tex}

\specID{SPEC-TH-041}
\specTitle{Field-Correlation Cortex --- real-time OSINT $\times$ Field corroboration}
\specStatus{DRAFT}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{Context}
The OSINT nano-swarm emits hypothesis edges at scale from open sources --- noisy, high-volume, low-trust. A separate high-trust field stream carries events from a tactical data-link, a sensor-fusion bus, friendly-force tracking, mission debriefs, and C2 feeds --- authoritative, low-volume, restricted. The value of the system is in fusing the two: OSINT raises hypotheses, field events confirm or refute them, and the knowledge graph reflects the fusion in quasi-real-time. This spec defines the \textbf{field-correlation cortex}, a new cortex subscribed to the tactical message bus (abstract over Kafka, ZeroMQ, MQTT) that consumes incoming field events, finds matching edges, and invokes \texttt{ConfidenceService} (SPEC-TH-040) to promote or demote them under a strict latency budget.

\section{Goals}
\begin{itemize}[leftmargin=1.2em]
  \item A new cortex \texttt{field-correlation} implementing the \texttt{Cortex} interface, registered alongside the other cortices.
  \item Subscribes to a tactical message bus behind a transport-agnostic \texttt{FieldStreamSubscriber} interface (Kafka, ZeroMQ, MQTT adapters).
  \item Matches incoming field events to candidate edges via \texttt{(subject, relation, object)} keying with configurable fuzziness (entity aliases, temporal windows).
  \item Invokes \texttt{ConfidenceService.promote} or \texttt{demote} on matches. No direct writes to the edge table --- always through the service (SPEC-TH-040 I-5).
  \item Meets a per-event latency budget: \textbf{$<$ 500 ms end-to-end} for \emph{critical} events, from bus arrival to provenance event emission.
  \item Emits per-event metrics (\texttt{match\_hit}, \texttt{match\_miss}, \texttt{contradiction}, \texttt{latency\_ms}) for observability.
\end{itemize}

\section{Non-Goals}
\begin{nongoal}
This spec does \emph{not} define the wire-level payload of the tactical data-link, the authentication / encryption of the bus, or the classification-level controls on field ingestion --- those live in deployment-side documents. It also does \emph{not} re-specify the confidence model; that is SPEC-TH-040. Multi-hop inference (e.g.\ promoting edges two hops away from a field-matched edge) is out of scope for v0.1.
\end{nongoal}

\section{Invariants}

\begin{invariant}
\textbf{I-1 --- Service-only writes.} The cortex never mutates edge rows directly. Every promotion or demotion goes through \texttt{ConfidenceService}, which enforces the per-class invariants from SPEC-TH-040.
\end{invariant}

\begin{invariant}
\textbf{I-2 --- Field dominance.} A field event always overrides OSINT confidence on a matching edge. The cortex must \emph{not} skip a promotion or demotion because of an older OSINT signal; field wins.
\end{invariant}

\begin{invariant}
\textbf{I-3 --- OSINT cannot demote FIELD\_HIGH.} The field-correlation cortex rejects and logs any request --- internal or upstream --- that would demote \texttt{FIELD\_HIGH} from a non-field actor.
\end{invariant}

\begin{invariant}
\textbf{I-4 --- Latency budget enforcement.} Every event carries a deadline derived from its \texttt{priority} ($\texttt{critical}$, $\texttt{routine}$, $\texttt{background}$). A processing path that exceeds its deadline emits a \texttt{LatencyBreach} event and continues; it does \emph{not} silently drop the event, and it does \emph{not} block the bus subscriber.
\end{invariant}

\begin{invariant}
\textbf{I-5 --- At-least-once correlation.} The cortex processes each field event at least once. It is idempotent under replay: re-applying the same event to the same edge is a no-op in terms of class and confidence value.
\end{invariant}

\begin{invariant}
\textbf{I-6 --- No cross-classification leakage.} Field events remain on their stream partition. A high-classification field event never produces a log line, metric label, or provenance field value that leaks content to a lower-classification observability sink. (Implementation: stream-tagged sinks; enforced by policy, asserted in tests.)
\end{invariant}

\section{Latency SLO table}

Per-event deadline from bus arrival to \texttt{EdgeProvenanceEvent} emission (p99 targets).

\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Priority} & \textbf{Examples} & \textbf{p50} & \textbf{p99 deadline} \\
\midrule
\texttt{critical}   & Engagement report, IFF anomaly, imminent-threat friendly-force track & 150 ms & \textbf{500 ms} \\
\texttt{routine}    & Track update, EW contact, routine C2 status change                  & 500 ms & \textbf{2{,}000 ms} \\
\texttt{background} & Post-mission debrief ingest, batched sensor-fusion replay           & 3 s    & \textbf{10 s} \\
\bottomrule
\end{tabular}
\end{center}

Budget is split approximately as: bus deserialize $\leq 10\%$, candidate-edge lookup $\leq 40\%$, confidence service call $\leq 30\%$, provenance emission $\leq 20\%$.

\section{Interface}

\begin{lstlisting}[language=TypeScript]
// packages/thalamus/src/cortices/field-correlation/types.ts

import type { Cortex, CortexInput, CortexOutput } from "../types";
import type { ConfidenceService, SourceClass } from "../confidence";

export type FieldStream =
  | "tactical-data-link"
  | "sensor-fusion-bus"
  | "friendly-force-tracking"
  | "c2-feed"
  | "mission-debrief"
  | "ew-contact"
  | "tactical-sigint";

export type FieldEventPriority = "critical" | "routine" | "background";

export interface FieldEvent {
  id: string;
  stream: FieldStream;
  priority: FieldEventPriority;
  receivedAt: Date;
  subject: { type: string; id: string; aliases?: string[] };
  relation: string;
  object:  { type: string; id: string; aliases?: string[] };
  outcome: "confirms" | "contradicts" | "partial";
  // Opaque payload retained for audit; never logged at lower classification.
  payload: Record<string, unknown>;
}

export interface FieldStreamSubscriber {
  subscribe(
    streams: FieldStream[],
    onEvent: (event: FieldEvent) => Promise<void>,
  ): Promise<void>;
  close(): Promise<void>;
}

export interface CorrelationResult {
  eventId: string;
  matchedEdgeIds: number[];
  outcome:
    | { kind: "promoted"; to: SourceClass }
    | { kind: "demoted";  to: SourceClass }
    | { kind: "no-match" }
    | { kind: "latency-breach"; budgetMs: number; actualMs: number };
  latencyMs: number;
}

export interface FieldCorrelationCortex extends Cortex {
  name: "field-correlation";
  // Streaming entry point; orchestrator starts this at boot.
  start(subscriber: FieldStreamSubscriber): Promise<void>;
  stop(): Promise<void>;
  // Cortex.execute is reused for replay / back-fill batches.
  execute(input: CortexInput): Promise<CortexOutput>;
}
\end{lstlisting}

\section{Registration}

The cortex is registered in \texttt{CortexRegistry} by a \texttt{field-correlation.md} skill file declaring its name, description, and the SQL helper used for candidate-edge lookup. At boot, the orchestrator resolves \texttt{FieldStreamSubscriber} from DI (bus transport swappable) and calls \texttt{cortex.start(subscriber)}.

\section{Scenarios}

\begin{scenario}{Critical field event confirms an OSINT hypothesis}
\Given an \texttt{OSINT\_CORROBORATED} edge at \texttt{confidence = 0.55} \\
\When a \texttt{critical} friendly-force tracking event arrives on the bus confirming the triple \\
\Then the cortex calls \texttt{ConfidenceService.promote} with \texttt{field-match} evidence \\
\And the edge becomes \texttt{FIELD\_HIGH} \\
\And end-to-end latency is below 500 ms (p99) \\
\And a \texttt{correlation.match\_hit} metric is emitted with \texttt{priority=critical}
\end{scenario}

\begin{scenario}{Routine field event with partial match}
\Given an \texttt{OSINT\_UNCORROBORATED} edge at \texttt{confidence = 0.30} \\
\When a \texttt{routine} sensor-fusion event arrives with outcome \texttt{partial} \\
\Then the edge is promoted to \texttt{FIELD\_LOW} \\
\And \texttt{confidence} is set to \texttt{0.75} \\
\And end-to-end latency is below 2{,}000 ms
\end{scenario}

\begin{scenario}{Field event contradicts a FIELD\_LOW edge}
\Given a \texttt{FIELD\_LOW} edge backed by an older field event \\
\When a contradicting field event arrives on the same stream \\
\Then the cortex calls \texttt{ConfidenceService.demote} with \texttt{field-contradiction} evidence \\
\And the edge drops to \texttt{OSINT\_UNCORROBORATED} with \texttt{confidence} $\leq 0.20$ \\
\And the edge is flagged for Sweep analyst review
\end{scenario}

\begin{scenario}{No matching edge}
\Given an incoming field event whose triple does not match any existing edge \\
\When the cortex runs the candidate-edge lookup \\
\Then no mutation is performed \\
\And a \texttt{correlation.match\_miss} metric is emitted \\
\And the event is retained on an unmatched-events queue for later correlation (bounded, TTL-gated)
\end{scenario}

\begin{scenario}{Latency breach on critical event}
\Given a \texttt{critical} event with budget 500 ms \\
\When the candidate-edge lookup exceeds the budget due to DB contention \\
\Then the cortex still completes the promotion \\
\And emits a \texttt{LatencyBreach} event with \texttt{budgetMs=500} and actual duration \\
\And the observability bus records an SLO violation \\
\And the event is not dropped
\end{scenario}

\begin{scenario}{OSINT attempts to demote FIELD\_HIGH}
\Given a \texttt{FIELD\_HIGH} edge \\
\When a late-arriving OSINT corroboration contradicts the field-confirmed relation \\
\Then the cortex refuses the demotion \\
\And logs a \texttt{rejected-demotion} event with actor \texttt{osint-swarm} \\
\And the edge class and confidence remain unchanged
\end{scenario}

\section{Acceptance Criteria}

\begin{ac}{AC-1}
The cortex registers under name \texttt{field-correlation} and is discoverable via \texttt{CortexRegistry.names()}.
\end{ac}

\begin{ac}{AC-2}
Starting the cortex with a mock \texttt{FieldStreamSubscriber} produces a delivered \texttt{critical} event routed through \texttt{ConfidenceService.promote}, ending in a \texttt{FIELD\_HIGH} edge within 500 ms (p99 over 1{,}000 synthetic events).
\end{ac}

\begin{ac}{AC-3}
A \texttt{routine} event completes end-to-end within 2{,}000 ms (p99 over 1{,}000 synthetic events); \texttt{background} within 10 s.
\end{ac}

\begin{ac}{AC-4}
A contradicting field event demotes a matching \texttt{FIELD\_*} edge to \texttt{OSINT\_UNCORROBORATED} with \texttt{confidence} $\leq 0.20$.
\end{ac}

\begin{ac}{AC-5}
A request that would demote \texttt{FIELD\_HIGH} from a non-field actor is rejected; edge state is unchanged; a \texttt{rejected-demotion} metric is emitted.
\end{ac}

\begin{ac}{AC-6}
Replaying the same \texttt{FieldEvent.id} produces no additional class or confidence change (idempotence under at-least-once delivery).
\end{ac}

\begin{ac}{AC-7}
An event with no matching edge emits \texttt{correlation.match\_miss} and persists to the bounded unmatched-events queue with a TTL.
\end{ac}

\begin{ac}{AC-8}
Every processed event emits an \texttt{EdgeProvenanceEvent} via \texttt{ConfidenceService} (when it mutates) with \texttt{actor = field-correlation}.
\end{ac}

\begin{ac}{AC-9}
A latency breach emits a \texttt{LatencyBreach} event tagged with \texttt{priority} and does not cause event loss.
\end{ac}

\begin{ac}{AC-10}
Observability sinks at lower classification never receive \texttt{FieldEvent.payload} content (asserted by a leakage test that inspects all emitted log lines and metric labels against an allow-list).
\end{ac}

\section{Traceability}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
AC & Test file (planned) & Test name \\
\midrule
AC-1  & \texttt{packages/thalamus/tests/field-correlation/registry.spec.ts}         & \texttt{field-correlation registers} \\
AC-2  & \texttt{packages/thalamus/tests/field-correlation/latency-critical.spec.ts} & \texttt{p99 $<$ 500ms on critical} \\
AC-3  & \texttt{packages/thalamus/tests/field-correlation/latency-routine.spec.ts}  & \texttt{p99 $<$ 2s routine / $<$ 10s background} \\
AC-4  & \texttt{packages/thalamus/tests/field-correlation/demote.spec.ts}           & \texttt{field contradiction demotes} \\
AC-5  & \texttt{packages/thalamus/tests/field-correlation/invariants.spec.ts}       & \texttt{OSINT cannot demote FIELD\_HIGH} \\
AC-6  & \texttt{packages/thalamus/tests/field-correlation/idempotence.spec.ts}      & \texttt{replay is no-op} \\
AC-7  & \texttt{packages/thalamus/tests/field-correlation/no-match.spec.ts}         & \texttt{miss routes to unmatched queue} \\
AC-8  & \texttt{packages/thalamus/tests/field-correlation/provenance.spec.ts}       & \texttt{provenance actor is field-correlation} \\
AC-9  & \texttt{packages/thalamus/tests/field-correlation/latency-breach.spec.ts}   & \texttt{breach emits event without drop} \\
AC-10 & \texttt{packages/thalamus/tests/field-correlation/leakage.spec.ts}          & \texttt{payload not in low-class sinks} \\
\bottomrule
\end{tabular}
\end{center}

\section{Open Questions}
\begin{itemize}[leftmargin=1.2em]
  \item Candidate-edge lookup: pgvector similarity on aliases vs. exact-key hash index? Likely hybrid --- exact first, vector fallback.
  \item Unmatched-events queue: Redis stream vs. a dedicated table with a TTL sweep? Leaning Redis stream with a length cap.
  \item Per-stream priority override: can a \texttt{mission-debrief} event ever be \texttt{critical}? Current draft: priority is event-level, stream is metadata.
  \item Back-pressure policy when bus rate exceeds worker capacity: shed \texttt{background} first, then \texttt{routine}; never \texttt{critical}.
  \item Classification policy for the unmatched-events queue: per-stream partition keys; test under I-6.
\end{itemize}

\end{document}
