% SPEC-TH-030 — Knowledge Graph Write-Path

\input{../preamble.tex}

\specID{SPEC-TH-030}
\specTitle{Knowledge Graph Write-Path --- structured entity persistence with provenance}
\specStatus{DRAFT}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{Context}
Thalamus cortices produce structured findings (entities, edges, evidence) that must land in
a Postgres + pgvector knowledge graph. Without a disciplined write-path, each cortex would
craft its own SQL, forget to stamp provenance, or drop evidence silently. This spec defines
the single write-path: a typed repository layer that every cortex is required to use, with
provenance enforced at the type system and at the database layer.

\section{Goals}
\begin{itemize}[leftmargin=1.2em]
  \item Every entity, edge, and finding reaches the graph through a typed repository method.
  \item Every row carries provenance: source identifier, cortex id, run id, UTC timestamp.
  \item Cortex code never composes SQL strings; the repository API is the only surface.
  \item Writes are idempotent via a \texttt{dedup\_hash} so retries do not duplicate findings.
\end{itemize}

\section{Non-Goals}
\begin{nongoal}
This spec does not cover the read-path (search, vector retrieval, DAG aggregation), schema
migrations, or the prompt templates that produce findings. Prompt authoring is covered in
SPEC-TH-031. Schema evolution is handled by the migrations stream.
\end{nongoal}

\section{Invariants}
\begin{invariant}
The agent MUST NOT issue ad-hoc SQL. All writes MUST flow through
\texttt{KnowledgeGraphRepository} methods. A lint rule forbids \texttt{pool.query},
\texttt{client.query}, and raw template literals containing \texttt{INSERT}, \texttt{UPDATE},
or \texttt{DELETE} anywhere under \texttt{packages/thalamus/src/cortices/}.
\end{invariant}

\begin{invariant}
Every persisted row carries non-null provenance: \texttt{(sourceUri, cortexId, runId,
createdAt)}. Inserts missing any provenance field are rejected at compile time (required
fields in the DTO) and at runtime (NOT NULL + CHECK constraints).
\end{invariant}

\begin{invariant}
Writes are idempotent. Replaying a batch with identical
\texttt{(cortexId, dedupHash)} leaves the store unchanged except for an incremented
\texttt{seenCount} and an updated \texttt{lastSeenAt}.
\end{invariant}

\section{Interface}

\begin{lstlisting}[language=TypeScript]
// packages/thalamus/src/cortices/storage/repository.ts

export interface Provenance {
  sourceUri: string;     // URL, DOI, or internal:<table>:<id>
  cortexId: string;      // e.g. "research-loop", "strategist"
  runId: string;         // ULID of the cortex cycle
  createdAt: Date;       // UTC
}

export interface EntityInput {
  type: string;          // "appellation" | "producer" | "vintage" | ...
  externalId: string;    // stable key in the source domain
  attributes: Record<string, unknown>;
  embedding?: number[];  // pgvector, optional
}

export interface EdgeInput {
  fromType: string; fromId: string;
  toType: string;   toId: string;
  relation: string;      // "about" | "produces" | "located_in" | ...
  weight: number;        // 0..1
}

export interface FindingInput {
  title: string;
  summary: string;
  findingType: "anomaly" | "trend" | "forecast"
    | "insight" | "alert" | "opportunity" | "strategy";
  confidence: number;    // 0..1
  impactScore: number;   // 0..10
  dedupHash: string;     // sha256(title + entityKey)
  evidence: Array<{ source: string; weight: number }>;
  edges: EdgeInput[];
}

export interface KnowledgeGraphRepository {
  upsertEntity(input: EntityInput, prov: Provenance): Promise<string>;
  upsertEdge(input: EdgeInput, prov: Provenance): Promise<void>;
  writeFinding(input: FindingInput, prov: Provenance): Promise<{
    id: string; deduplicated: boolean;
  }>;
  writeFindings(batch: FindingInput[], prov: Provenance): Promise<{
    inserted: number; deduplicated: number;
  }>;
}
\end{lstlisting}

\section{Scenarios}

\begin{scenario}{Nominal write --- new finding with entity and edge}
\Given a cortex run with \texttt{runId = R1} and \texttt{cortexId = value-detective} \\
\When \texttt{writeFinding} is called with a fresh \texttt{dedupHash} referencing entity
\texttt{appellation:pauillac} \\
\Then a row is inserted in \texttt{finding} with provenance \texttt{(R1, value-detective)} \\
\And referenced entities and edges are upserted in the same transaction \\
\And the returned \texttt{deduplicated} flag is \texttt{false}.
\end{scenario}

\begin{scenario}{Idempotent replay}
\Given a finding with \texttt{dedupHash = H} has already been persisted by cortex \texttt{C} \\
\When the same batch is replayed after a crash \\
\Then no new row is created \\
\And the existing row's \texttt{seenCount} is incremented and \texttt{lastSeenAt} bumped \\
\And the returned \texttt{deduplicated} flag is \texttt{true}.
\end{scenario}

\begin{scenario}{Missing provenance is refused}
\Given a caller constructs a \texttt{FindingInput} but omits \texttt{cortexId} from \texttt{Provenance} \\
\When the code is compiled \\
\Then the TypeScript compiler rejects the call (required field) \\
\And the corresponding runtime \texttt{CHECK} constraint would also reject a direct insert.
\end{scenario}

\begin{scenario}{Raw SQL in a cortex is blocked at review}
\Given a contributor adds \texttt{await pool.query("INSERT INTO finding ...")} in a cortex file \\
\When the lint step runs in CI \\
\Then the build fails with rule \texttt{no-raw-sql-in-cortices} \\
\And the PR cannot be merged until the call is routed through the repository.
\end{scenario}

\begin{scenario}{Partial failure is atomic}
\Given a finding references three edges and the third edge violates a FK constraint \\
\When \texttt{writeFinding} runs \\
\Then the enclosing transaction is rolled back \\
\And neither the finding nor any of its edges are visible to readers.
\end{scenario}

\section{Acceptance Criteria}

\begin{ac}{AC-1}
Static analysis (lint rule \texttt{no-raw-sql-in-cortices}) reports zero violations across
\texttt{packages/thalamus/src/cortices/} on every PR. A seeded violation fails CI.
\end{ac}

\begin{ac}{AC-2}
\texttt{writeFinding} called twice with the same \texttt{(cortexId, dedupHash)} results in
exactly one row in \texttt{finding}; the second call returns \texttt{deduplicated = true}.
\end{ac}

\begin{ac}{AC-3}
Inserting a row into \texttt{finding}, \texttt{entity}, or \texttt{edge} with any of
\texttt{sourceUri, cortexId, runId, createdAt} set to \texttt{NULL} is rejected by the
database.
\end{ac}

\begin{ac}{AC-4}
A finding whose edge list contains an invalid reference causes the full transaction to roll
back; a post-test SELECT shows no partial state.
\end{ac}

\begin{ac}{AC-5}
The repository exposes only the four methods listed in the Interface section; no other
public export touches the database.
\end{ac}

\begin{ac}{AC-6}
Every row in \texttt{finding} produced during an end-to-end cortex cycle can be linked back
to its originating \texttt{runId} via a simple \texttt{SELECT}.
\end{ac}

\section{Traceability}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
AC & Test file & Test name \\
\midrule
AC-1 & \texttt{packages/thalamus/tests/lint/no-raw-sql.spec.ts} & \texttt{rejects raw INSERT in cortex} \\
AC-2 & \texttt{packages/thalamus/tests/storage/repository.spec.ts} & \texttt{dedup replay is idempotent} \\
AC-3 & \texttt{packages/thalamus/tests/storage/repository.spec.ts} & \texttt{missing provenance is rejected} \\
AC-4 & \texttt{packages/thalamus/tests/storage/repository.spec.ts} & \texttt{partial failure rolls back} \\
AC-5 & \texttt{packages/thalamus/tests/storage/repository-api.spec.ts} & \texttt{public surface is frozen} \\
AC-6 & \texttt{packages/thalamus/tests/integration/cycle-provenance.spec.ts} & \texttt{findings link to runId} \\
\bottomrule
\end{tabular}
\end{center}

\section{Open Questions}
\begin{itemize}[leftmargin=1.2em]
  \item Should \texttt{embedding} be written in the same transaction as the entity or in a
        follow-up async job? Latency vs. atomicity trade-off.
  \item Do we need a soft-delete flag on \texttt{finding} for retraction, or is a new
        finding with \texttt{findingType = "retraction"} enough?
  \item Per-cortex rate limits on writes --- enforced here or upstream in the executor?
\end{itemize}

\end{document}
