\input{../preamble.tex}

\specID{SPEC-TH-011}
\specTitle{Source Fetchers --- typed adapter contract for external knowledge sources}
\specStatus{DRAFT}
\specVersion{0.1}
\specOwner{Jeremie Nunez}

\begin{document}
\makespecheader

\section{Context}
Thalamus cortices enrich their reasoning with data from heterogeneous external sources
(climate archives, terroir databases, market feeds, regulatory corpora, knowledge graphs, \ldots).
Each of these sources speaks its own protocol and returns its own shape. To prevent the complexity
from leaking into cortex logic, the system defines a single typed adapter contract --- the
\texttt{SourceFetcher} --- and a registry that routes cortex requests to the right set of fetchers.
This spec fixes that contract.

\section{Goals}
\begin{itemize}[leftmargin=1.2em]
  \item Define one canonical \texttt{SourceFetcher} function type that all adapters must implement.
  \item Define a uniform \texttt{SourceResult} envelope so consumers can treat heterogeneous sources homogeneously.
  \item Provide a registry that maps cortex names to an ordered list of fetchers, supports self-registration, and caches results per \((cortex, params)\) key.
  \item Guarantee that a single fetcher failure never cancels the other fetchers of the same cortex.
\end{itemize}

\section{Non-Goals}
\begin{nongoal}
This spec does not cover: the internal implementation of any specific fetcher, the schema of the
\texttt{data} payload (opaque to the registry), retry and circuit-breaker policy inside a fetcher,
and cortex-level fusion of fetcher results (downstream of this contract).
\end{nongoal}

\section{Invariants}
\begin{invariant}
A \texttt{SourceFetcher} is a pure async function from an untyped parameter bag to a
\emph{list} of \texttt{SourceResult}. It may return an empty list but must never return \texttt{undefined} or \texttt{null}.
\end{invariant}

\begin{invariant}
Every \texttt{SourceResult} carries its own \texttt{type}, \texttt{source}, \texttt{fetchedAt} ISO timestamp,
and \texttt{latencyMs}. These fields are non-optional.
\end{invariant}

\begin{invariant}
\texttt{fetchSourcesForCortex} is non-throwing: any rejection from an underlying fetcher is absorbed
(logged, not propagated) and only the successful fetchers contribute to the returned list.
\end{invariant}

\begin{invariant}
Results are cached per \((cortex, JSON(params))\) key for \texttt{CACHE\_TTL\_MS}; cache hits return
the exact same \texttt{SourceResult[]} reference shape as a live fetch.
\end{invariant}

\begin{invariant}
Registration is additive: \texttt{registerSource} can be called multiple times for the same cortex
and preserves insertion order; it never removes a previously registered fetcher.
\end{invariant}

\section{Interface}
The contract is deliberately narrow: one function type, one envelope type, two registry operations.

\begin{lstlisting}[language=TypeScript]
export interface SourceResult {
  type: string;         // e.g. "climate.huglin", "market.index", "regulation.doc"
  source: string;       // human-readable origin, e.g. "open-meteo", "knowledge-graph"
  url?: string;         // optional deep link to the original record
  data: unknown;        // opaque payload; shape is fetcher-specific
  fetchedAt: string;    // ISO-8601 timestamp, UTC
  latencyMs: number;    // wall-clock latency of the fetch
}

export type SourceFetcher = (
  params: Record<string, unknown>,
) => Promise<SourceResult[]>;

export function registerSource(
  cortices: string[],
  fetcher: SourceFetcher,
): void;

export function fetchSourcesForCortex(
  cortexName: string,
  params: Record<string, unknown>,
): Promise<SourceResult[]>;
\end{lstlisting}

\section{Scenarios}

\begin{scenario}{Nominal multi-fetcher cortex}
\Given a cortex \texttt{C} has three fetchers registered, all of which succeed \\
\When \texttt{fetchSourcesForCortex("C", params)} is called \\
\Then the returned list concatenates the results of all three fetchers, in registration order \\
\And every result carries \texttt{fetchedAt} and \texttt{latencyMs}
\end{scenario}

\begin{scenario}{One fetcher fails}
\Given a cortex has three fetchers and the second one throws \\
\When \texttt{fetchSourcesForCortex} is called \\
\Then the returned list contains only the results of the first and third fetchers \\
\And the call resolves successfully (never rejects) \\
\And the failure is logged but not re-raised
\end{scenario}

\begin{scenario}{Unknown cortex}
\Given no fetchers have been registered for cortex \texttt{"unknown"} \\
\When \texttt{fetchSourcesForCortex("unknown", anything)} is called \\
\Then the result is an empty array \\
\And no fetcher is invoked
\end{scenario}

\begin{scenario}{Cache hit}
\Given a cortex \texttt{C} was queried with params \texttt{P} less than \texttt{CACHE\_TTL\_MS} ago \\
\When \texttt{fetchSourcesForCortex("C", P)} is called again with identical params \\
\Then the cached list is returned \\
\And no underlying fetcher is invoked
\end{scenario}

\begin{scenario}{Cache miss on distinct params}
\Given the cache holds an entry for \texttt{("C", \{a:1\})} \\
\When \texttt{fetchSourcesForCortex("C", \{a:2\})} is called \\
\Then the cached entry is not used \\
\And all registered fetchers for \texttt{C} are invoked
\end{scenario}

\section{Acceptance Criteria}

\begin{ac}{AC-1}
\texttt{fetchSourcesForCortex} for a cortex with zero registered fetchers resolves to \texttt{[]} without invoking any side effect.
\end{ac}

\begin{ac}{AC-2}
A \texttt{SourceFetcher} that rejects with any error does not cause \texttt{fetchSourcesForCortex} to reject; the peer fetchers' results are still returned.
\end{ac}

\begin{ac}{AC-3}
Two consecutive calls with identical \texttt{(cortex, params)} within \texttt{CACHE\_TTL\_MS} invoke the underlying fetchers exactly once in total.
\end{ac}

\begin{ac}{AC-4}
Two calls with the same cortex but distinct \texttt{params} (deep-inequal by JSON) each invoke the underlying fetchers.
\end{ac}

\begin{ac}{AC-5}
Calling \texttt{registerSource(["A","B"], f)} makes \texttt{f} contribute to both \texttt{A} and \texttt{B} results.
\end{ac}

\begin{ac}{AC-6}
Every \texttt{SourceResult} returned by the registry has \texttt{type}, \texttt{source}, \texttt{fetchedAt} (valid ISO-8601), and \texttt{latencyMs \(\ge 0\)} present.
\end{ac}

\begin{ac}{AC-7}
The function type \texttt{SourceFetcher} has exactly one arity (a single \texttt{params} argument) and always returns \texttt{Promise<SourceResult[]>} --- a type-level test rejects any narrower or wider shape.
\end{ac}

\section{Traceability}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
AC & Test file & Test name \\
\midrule
AC-1 & \texttt{packages/thalamus/tests/source-registry.unknown.spec.ts} & \texttt{returns empty for unregistered cortex} \\
AC-2 & \texttt{packages/thalamus/tests/source-registry.failure.spec.ts} & \texttt{absorbs single fetcher rejection} \\
AC-3 & \texttt{packages/thalamus/tests/source-registry.cache.spec.ts} & \texttt{caches identical queries} \\
AC-4 & \texttt{packages/thalamus/tests/source-registry.cache.spec.ts} & \texttt{bypasses cache for distinct params} \\
AC-5 & \texttt{packages/thalamus/tests/source-registry.register.spec.ts} & \texttt{registers across multiple cortices} \\
AC-6 & \texttt{packages/thalamus/tests/source-registry.envelope.spec.ts} & \texttt{enforces SourceResult envelope fields} \\
AC-7 & \texttt{packages/thalamus/tests/source-fetcher.types.spec-d.ts} & \texttt{SourceFetcher type is exact} \\
\bottomrule
\end{tabular}
\end{center}

\section{Open Questions}
\begin{itemize}[leftmargin=1.2em]
  \item Should the registry support cache invalidation per-cortex or per-key (currently TTL only)?
  \item Should \texttt{params} be typed per cortex (generic \texttt{SourceFetcher<P>}) rather than \texttt{Record<string, unknown>}?
  \item Should failing fetchers be surfaced to the caller as a diagnostic channel, or remain logs-only?
\end{itemize}

\end{document}
